Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-04-24 Manuscript/manuscript.tex
Still more introduction edits
Restructuring, streamlining, and adding a few citations to first
paragraph to justify importance of env. transmission.

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.  This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infected individual.  This additional transmission pathway can have important consequences for disease dynamics, especially in spatially structured populations (\cite{Park2012}).

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect residents of the local patch), its role in disease dynamics may depend on the attributes of the specific habitat patches in which reservoirs become established.  In particular, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  Thus, we expect that patterns of host movement, and consequently the effect of environmental pathogen transmission, are determined at least in part by patch quality.

Many of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies), and thus understanding how environmental pathogen persistence interacts with habitat quality is critical to managing disease in these systems.  In this study, we ask how pathogen longevity (how long the pathogen can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality across the metapopulation) interact to influence pathogen spread and metapopulation occupancy.  Within a metapopulation, we ask what specific role low and high quality patches play in controlling pathogen spread.

We expect pathogen spread to increase with both environmental longevity and with habitat heterogeneity.  Increased environmental longevity should provide greater opportunities for transmission, and thus should favor pathogen spread.  The role of habitat heterogeneity is likely to be more complex, emerging from the behavior of high and low quality patchs in heterogeneous metapopulations.  Specifically, we expect that high quality habitat patches, which support greater host traffic than lower quality habitat, might be more likely to form pathogen reservoirs.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}, citations, examples?).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).  Similarly, we predict that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and and thus may serve as refuges for susceptible hosts to escape infection.

To address the above questions and evaluate our hypotheses, we build on the metapopulation framework developed by \cite{Hanski1994} to develop and explore a theoretical stochastic patch occupancy model that incorporates pathogen transmission from environmental reservoirs and habitat heterogeneity.

@@

>>>>

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.  This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infected individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}.

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.  In particular, we expect that the host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.  Indeed, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes in turn may then influence where environmental pathogen reservoirs get established and how they affect disease dynamics and host occupancy throughout the metapopulation.

Specifically, we expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).  Similarly, we expect that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and thus may serve as refuges on which susceptible hosts can escape infection.

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality in a metapopulation) interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.

@@

Also discuss \cite{Sharp2011}, who show that management actions that increase carrying capacity can increase disease spread and destabilize population.  And \cite{Hess1996} who finds that implementing a quarantine on centralized patches reduces pathogen spread.
